---
title: "a04"
author: "Patrick Sherry"
date: "3/26/2026"
format: html
---
## Introduction

This report practices joining tables together using bird survey data collected from the central Arizona-Phoenix metropolitan area by Arizona State University researchers.

## Analysis

```{r}
library(readr)
library(dplyr)
library(here)
```

```{r}

# read in data
bird_observations <- readr::read_csv(here::here("data", "raw", "52_pp52_birds_1.csv"))
surveys <- readr::read_csv(here::here("data", "raw", "52_pp52_surveys_1.csv"))
sites <- readr::read_csv(here::here("data", "raw", "52_pp52_sites_1.csv"))
taxalist <- readr::read_csv(here::here("data", "raw", "52_pp52_taxalist_1.csv"))
```

```{r}
# dimensions of the dataframe
dim(bird_observations)

# returns top six rows
head(bird_observations)

# returns all the columns
glimpse(bird_observations)

# returns some statistics
summary(bird_observations)

# returns column names
names(bird_observations)

# returns unique values
unique(bird_observations$species_id)
```

```{r}
# create a subset of bird_observations with only BHCO and RWBL species
birds_subset <- bird_observations %>%
  filter(species_id %in% c("BHCO", "RWBL")) %>%
  filter(site_id %in% c("LI-W", "NU-C"))
```

```{r}
# use a left join to merge birds_subset with sites
birds_left <- birds_subset %>%
  left_join(y = sites, by = "site_id")
```

```{r}
# use a full join to merge birds_subset and sites
birds_full <- birds_subset %>%
  full_join(y = sites, by = "site_id")
```

```{r}
# use an inner join to merge birds_subset and taxalist
birds_inner <- birds_subset %>%
  inner_join(y = taxalist, by = "species_id")
```

```{r}
# use a left join instead of inner join for comparison
birds_inner_left <- birds_subset %>%
  left_join(y = taxalist, by = "species_id")
```

```{r}
# rename the notes column to bird_obs_notes
bird_observations <- bird_observations %>%
  rename(bird_obs_notes = notes)
```

```{r}
# join all tables together and select only the columns needed
bird_obs_subset <- bird_observations %>%
  full_join(y = taxalist, by = "species_id") %>%
  full_join(y = sites, by = "site_id") %>%
  full_join(y = surveys, by = c("site_id", "survey_id")) %>%
  select(survey_date, common_name, park_name, bird_count, observer)
```

```{r}
# check the unique observer values
unique(bird_obs_subset$observer)
```

```{r}
# fix "Josh Burns" to match the format of other observer names
bird_obs_subset <- bird_obs_subset %>%
  mutate(observer = if_else(condition = observer == "Josh Burns",
                            true = "J. Burns",
                            false = observer))

unique(bird_obs_subset$observer)
```

## Discussion

Joining the four bird survey tables together allowed us to connect species names, site information, and survey metadata to the raw bird observations. The data cleaning revealed an inconsistently formatted observer name which was corrected to match the format used by others.
